Revisiting~the~Classics: Online~RL~in~the~Programmable~Dataplane

### 2021-06-11 paper/paper-10pt.tex
Before ImageUp

%\vspace{-1em}
\section{Introduction}
\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.65\linewidth]{figures/arch-with-p4}
	\caption{\approachshort{} brings low-latency, online reinforcement learning directly to the dataplane. SoC- and NetFPGA-based SmartNIC devices expose spare compute---making in-situ, asynchronous processing and learning possible alongside P4 dataplanes. Classical RL policy methods are the key to making this computationally feasible.\label{fig:netro-arch}}
\end{figure}
Automatic network optimisation, control, and defence are at last becoming commonplace.
Adaptive techniques such as \emph{reinforcement learning} (RL) have led the charge in data-driven networking, enhancing automatic traffic optimisation~\parencite{DBLP:conf/sigcomm/ChenL0L18,DBLP:conf/sigcomm/MaoNA17}, congestion control~\parencite{DBLP:journals/corr/abs-1910-04054}, adaptive routing~\parencite{DBLP:conf/hotnets/ValadarskySST17,DBLP:conf/conext/GiladSGRS20}, resource management~\parencite{DBLP:conf/hotnets/MaoAMK16}, and packet classification~\parencite{DBLP:conf/sigcomm/LiangZJS19}.
In RL methods, every change and its effects improve future decisions made by an agent.

@@

In this paper, we enable \emph{online in-NIC learning} by returning to \emph{classical} RL methods.
In particular, we focus on tile coding with one-step temporal-difference algorithms such as Sarsa~\parencite{RL2E}.
%These choices have important benefits for in-NIC execution.
These functions have lower model capacity, they do not require batches of inputs to learn in a stable way, negating the memory needed to store experience replays, and have simple update and inference logic.
In addition, they have shown promise in other network use cases~\parencite{DBLP:journals/eaai/MalialisK15,DBLP:journals/tnsm/SimpsonRP20}.
Tile coding in particular admits many optimisations, being an embarrassingly parallel problem.
Using fixed-point arithmetic, we solve the lack of floating-point support in PDP hardware \emph{and} enable new optimisations.
Moreover, the P4 dataplane can offer runtime control over which flows are monitored.
%Finally, the choice of single-step algorithms (as opposed to $n$-step or Monte Carlo methods) bounds the amount of per-trace state required for online learning to just the last state-action pair, safeguarding the limited memory of the target devices.
We also design our solution to operate as closely as possible to the P4 pipeline to use and learn from per-packet state, but outside of the main packet path to prevent packet stalls.

@@

While P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to great interest in network programmability, it requires similar behaviour between device classes.
This is encoded by the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}---a lower bound on the device capabilities required to implement a P4 dataplane.
%We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
However, many compatible devices have legacies long predating these developments. 
For instance, many-core SOC-based Netronome~\parencite{netronome-smartnic}, NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14,DBLP:conf/fpga/IbanezBMZ19}, and other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo} allow virtually arbitrary programs to be specified and executed.
%This pattern extends to other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo}.
Currently, low port-density devices like these are most likely to have specific general-purpose compute, as they are designed for high-performance offloading and middlebox development.
%Even Intel Tofino ASICs~\parencite{barefoot-intel}, which architecturally mirror the PSA, expose additional matching capabilities and \emph{arithmetic logic unit} (ALU) functions via the Tofino Native Architecture.
Regardless, floating-point operations key to ML/RL workloads are very rarely supported outside of deep learning accelerators (which lack packet switching functionality).
Moreover, the limited memory and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

@@

\emph{Reinforcement learning} (RL) algorithms train an \emph{agent} to choose an optimal sequence of actions in pursuit of a given task~\parencite{RL2E}.
Like most ML methods, RL algorithms use gradient information to update the parameters used to approximate a function with \emph{reward measurements} from the environment.
RL's \emph{Markov Decision Process} (MDP) structure allows online learning of a state-action map in a model-free way, and can step through local minima when needed compared to ML.
In networking, this allows for learning to respond to on-device state or to handle rapidly evolving problems.
RL algorithms are generally agnostic to the policy approximation used, and classical single-step methods are simple---most of the compute work is occupied by the policy itself.
The single-step, semi-gradient Sarsa algorithm~\parencite[pp. \numrange{217}{221}]{RL2E} requires only additions and multiplication to update and learn a policy online.

@@

\subsection{Representations for embedded ML}
As PDP hardware universally lacks floating-point arithmetic support, the task of bringing ML models to resource-constrained environments without these capabilities is well-studied.
Quantisation and alternative data formats have been suggested to make ML inference feasible on resource- and power-limited platforms, work around hardware constraints, or compute faster and more efficiently.
Lower bit depths reduce memory footprints, and improve throughput in e.g., \emph{bfloat16}~\parencite{bfloat16-blog,DBLP:journals/sigops/XieDMKVZT18}, \emph{hfp8}~\parencite{DBLP:conf/nips/SunCCWVSCZG19}, and other floating point formats~\parencite{DBLP:journals/corr/abs-2007-01530}.
In many cases, accuracy losses are negligible.
Such work goes further still towards integers~\parencite{tensorrt-8bit} or binarisation~\parencite{DBLP:journals/corr/MiyashitaLM16,DBLP:conf/eccv/RastegariORF16,DBLP:journals/corr/KimS16,DBLP:conf/nips/HubaraCSEB16}, sacrificing dynamic range for cheaper arithmetic.
Similar techniques are used to make inference tractable on network hardware, i.e., to bring pre-trained neural networks to the dataplane~\parencite{DBLP:journals/corr/abs-2009-02353,DBLP:conf/sigcomm/SanvitoSB18,DBLP:journals/corr/abs-1801-05731}.

@@

\subsection{Action and Update Computation}\label{sec:action-and-update-computation}
\approachshort{} applies the insights of \textcite{DBLP:journals/firai/TravnikMSP18} to minimise action latency; an action is computed, sent out into the environment, and only then is the underlying policy updated.
Using one of the below strategies chosen at compile time, a state vector is tile coded, converted into action probabilities, and an action is chosen.
This is then written out to the environment as in \cref{sec:agent-environment-communication}.
If online learning is enabled, \approachshort{} then checks an internal hashmap for a previous state-action pair matching the current instruction source, and if found then the policy is updated.
Updates are computed using \emph{single-step semi-gradient Sarsa}~\cite[pp. \numrange{217}{221}]{RL2E}, though modification to support other single-step methods would be trivial.
The new state-action pair is then written into storage.
\approachshort{} can be configured to automatically select values of the input state vector as keys for state and reward storage.
Two firmware models govern how these tasks (action selection, policy updates) are carried out:
\begin{description}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each works sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate on processing state vectors, minimising latency. \emph{Minion} threads have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{description}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks to update the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput have different effects on RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem).

@@

\subsection{Agent-Environment Communication}\label{sec:agent-environment-communication}
\approachshort{} uses \emph{multiple-producer/multiple-consumer} (MPMC) channels to communicate with other elements; be they P4 programs on the packet path, or other on-chip analysis and control modules.
Through these, a system \emph{pushes} state vectors, reward measures, and setup packets as inputs, and \emph{pulls} a stream of state-action pairs as outputs.
This allows decisions to be made asynchronously, preventing packet stalling.
The key insight of this mechanism is that on-chip reward/state signals enjoy first-class support in the same manner as packets from the P4 dataplane, allowing agents to act on environmental signals from other on-NIC/chip asynchronous processes or the controller.
As such, \approachshort{} can receive input from P4 \texttt{extern}s or other, dedicated off-path flow state measurement applications.

Our implementation uses platform-specific IPC (\emph{EMEM ring buffers}) with hardware signalling on work arrival to achieve this.
As PDP hardware lacks dynamic memory allocation, P4 pipeline threads request buffers for packet payloads using a shared freelist to enable state/configuration/policy data handover.
Packet headers are extracted and parsed using the tooling autogenerated by the P4 pipeline.
We found that this costs a median \SIrange{126}{140}{\nano\second} communication time (local--remote), comparable to message channels in the Rust and Go languages on commodity hardware.

@@

Our implementation exploits the locality of cores in the NFP, which can be factored into designs on other platforms like NetFPGA as additional off-path functional units, as used by \textcite{DBLP:conf/isca/LiLYCSH19}.
Policy compute/update/configure tasks are passed between cores using direct \emph{next neighbour} registers, signalling all child threads in response.
Such on-chip signals cost just $\sim$\SI{20}{\nano\second} per relayed message.
Each core performs atomic adds to a shared preference list and an acknowledgement counter checked by the master thread, implementing our wait-free \emph{ParSa} algorithm (\cref{alg:parsa}).
This is essential for aggregation compared to the use of bounded message buffers, which caused significant head-of-line blocking.
%For instance, additional off-path functional units can serve as workers, like the floating-point adders used by \textcite{DBLP:conf/isca/LiLYCSH19}.
%Rather than general-purpose IPC, \approachshort{} would be hard-wired between internal workers and to the P4$\rightarrow$NetFPGA~\parencite{DBLP:conf/fpga/IbanezBMZ19} actions that interface with it.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and learning parameters to be changed at runtime using two control packets.
Design changes are used at the end of learning (moving from online to offline), or when trying to train a new policy for another task from the same vantage point.
Parameter changes allow an online agent to become more (or less) adaptive to new data (i.e., after detecting a changepoint in traffic).
This extends to policy data, which may be imported from a pre-trained model and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, \Coopfw/\Indfw, and maximum policy/tiling/state sizes---these govern core operation or allocated memory.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing more complex problems to be modelled using more dimensions or fine-grained tiles.

@@

Policy sizes are set to those of a real-world DDoS control application~\parencite{DBLP:journals/tnsm/SimpsonRP20}: 20-dim state vectors, a bias tile and 16 full tiling sets (\numproduct{7x1}-dim, \numproduct{8x2}-dim, \numproduct{1x4}-dim), 8 tilings per set, 6 tiles per dimension, and 10 actions.
For context, such input would contain per-flow state (e.g., IATs, rates) which is combined with the last action taken (2-dim tilings) and loads along the ingress-egress path (4-dim).
In \Coopfw{}, this creates \num{129} work items across \num{31} workers.
We choose a larger action than this work to investigate the performance of more complex agents.

@@

We then vary the amount of worker threads to show how \approachshort{} scales to fit available compute resources on a device.
This is an important aspect for (automated) allocation of compute resources in an intelligent dataplane---particularly when cohabiting with other dataplane programs---and has effects on ahead-of-time work scheduling which we examine later.
This also demonstrates the number of cores needed to achieve a given latency/throughput bound on a policy of representative complexity.
Moreover, to demonstrate how these costs vary as policy complexity increases, we vary the number of total dimensions included in the tiling.

@@

We perform these tests using Pktgen-DPDK~\cite{pktgen-dpdk}, placing an NFP in \emph{MidServer} as the device under test and connecting \emph{HighServer} over a \SI{40}{\giga\bit\per\second} DAC cable as the traffic source via the default firmware.
We perform throughput/loss tests using \num{7}/\num{1} transmit/receive queues at \SI{100}{\percent} send rate for 10 bursts of \SI{30}{\second}, and perform latency tests using \num{1}/\num{1} transmit/receive queue at \SI{10}{\percent} send rate for \num{200000} measurements (sampling at \SI{2000}{\hertz} for \qtyproduct[product-units=single]{10 x 10}{\second}).
This provides maximum throughput in the former case (relying solely on NIC counters for loss counting).
In the latter case, this minimises host resource contention to observe exact latency measurements, have a high enough sample count to detect subtle (aggregate) latency effects, and eliminate \emph{host} receive drops.
DPDK was setup using \qtyproduct[product-units=single]{4 x 1}{\gibi\byte} hugepages.
Sent traffic was comprised of fixed-size \SIrange{64}{1518}{\byte} packets~\cite{rfc2544}.
CPU clock scaling was disabled on \emph{HighServer}.

\fakepara{Resource requirements}
Using the maximum policy size defined above, we investigate how the memory requirements imposed by \approachshort{} vary with the number of dedicated MEs, over and above a base P4 forwarding plane.
We report resource use for \SI{32}{\bit} \Indfw{} and \Coopfw{} agents, with hash-tables sized to \num{4096} state-action pairs and \num{16} separate reward values.
This captures the relative cardinality of network RL traces to rewards, i.e., many input flows will map to one or few reward values (i.e., DDoS attack size estimation per egress-AS, queue occupancy in the case of AQM per output port).

@@

By limiting the available workers in software, we show how \Coopfw{}'s policy update time (thus  online throughput---\cref{fig:vary-core}) scales with available cores.
While \Coopfw{} always outperforms the host-based floating point implementations, we observe that there are two distinct crossover points which must be met to overcome our own \Indfw{}; \num{8} workers for online throughput, and \num{3} workers for state-action latency (plot omitted).
Some artefacts of our environment and design choices are visible, such as the addition of new physical cores being more significant than contexts, and the presence of some schedule bottlenecks.
Most importantly however, \Coopfw{}'s resource demand is tunable at compile time to meet the online training rate and/or action latency required by a task/environment.

@@

%A noteworthy trend is that \SIlist{8;16}{\bit} versions of \approachshort{} consistently underperform compared to \SI{32}{\bit}, except for smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency}).
\SIlist{8;16}{\bit} versions of \approachshort{} underperform compared to \SI{32}{\bit}, except for smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency})---even though our implementation is optimised to read and write policy data in batches.
Because the native register width on the NFP is \SI{32}{\bit}, the compiler must emit extra instructions around arithmetic operations to correctly load and update values.
This matches \SI{32}{\bit} becoming dominant in complex workloads: higher dimension tilings require more arithmetic operations.
Most of the I/O comes \emph{after} this step, causing ALU use to dominate.
This also explains why \SI{32}{\bit} becomes the best choice at different policy complexities for online (\cref{fig:vary-work}, 10 dims) and offline (\cref{fig:vary-work-latency}, 3 dims) agents, where hashtable accesses and a \texttt{memcpy} of the state vector fall into the serial portion of the online algorithm.
%Smaller bit depths still give a \SIrange{2}{4}{$\times$} saving in memory for policy storage (enabling more fine-grained or complex policies), in exchange for slightly worse latency and throughput.
To overcome this, we investigated bit-stuffing several values into a single word during atomic writeback (as the platform offers both \SIlist{32;64
}{\bit} atomic addition).
This is analogous to SIMD through clever use of padding bits, but we found that manipulating tiles into the correct format added \SI{10}{\percent} extra overhead.

@@

\fakepara{Resource requirements}
\Cref{tab:resources} shows how \approachshort{} consumes memory as it scales to fit a device's compute resources, compared with a simple P4 forwarding application.
As one program is installed per ME, these results represent the minimum and maximum resource use on a single island (i.e., without replacing P4 workers).
We observe negligible costs on shared EMEM ($\sim$\SI{4}{\mebi\byte}) from state/reward hashtables.
The most significant costs arise due to policy data (\SI{405}{\kibi\byte} shared IMEM, \SI{90}{\kibi\byte} local CTM, \SI{15}{\kibi\byte} local CLS), which can be halved or quartered using \SIlist{16;8}{\bit} quantisation.
This is a high upfront cost on per-island resources (CLS/CTM)---\approachshort{} leaves resources for other off-path dataplane applications, but is fairest from 3 cores onwards.

>>>>

%\vspace{-1em}
\section{Introduction}
\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.6875\linewidth]{figures/arch-with-p4}
	\caption{\approachshort{} brings low-latency, online reinforcement learning directly to the dataplane. SoC- and NetFPGA-based SmartNIC devices expose spare compute---making in-situ, asynchronous processing and learning possible alongside P4 dataplanes. Classical RL policy methods are the key to making this computationally feasible.\label{fig:netro-arch}}
\end{figure}
Automatic network optimisation, control, and defence are at last becoming commonplace.
Adaptive techniques such as \emph{reinforcement learning} (RL) have led the charge in data-driven networking, enhancing automatic traffic optimisation~\parencite{DBLP:conf/sigcomm/ChenL0L18,DBLP:conf/sigcomm/MaoNA17}, congestion control~\parencite{DBLP:journals/corr/abs-1910-04054}, adaptive routing~\parencite{DBLP:conf/hotnets/ValadarskySST17,DBLP:conf/conext/GiladSGRS20}, resource management~\parencite{DBLP:conf/hotnets/MaoAMK16}, and packet classification~\parencite{DBLP:conf/sigcomm/LiangZJS19}.
In RL methods, every change and its effects improve future decisions made by an agent.

@@

In this paper, we enable \emph{online in-NIC learning} by returning to \emph{classical} RL methods.
In particular, we focus on tile coding with one-step temporal-difference algorithms such as Sarsa~\parencite{RL2E}.
%These choices have important benefits for in-NIC execution.
While these functions have lower model capacity, they do not require batches of inputs to learn in a stable way, negating the memory needed to store experience replays, and have simple update and inference logic.
In addition, they have shown promise in other network use cases~\parencite{DBLP:journals/eaai/MalialisK15,DBLP:journals/tnsm/SimpsonRP20}.
Tile coding in particular admits many optimisations, being an embarrassingly parallel problem.
Using fixed-point arithmetic, we solve the lack of floating-point support in PDP hardware \emph{and} enable new optimisations.
Moreover, the P4 dataplane can offer runtime control over which flows are monitored.
%Finally, the choice of single-step algorithms (as opposed to $n$-step or Monte Carlo methods) bounds the amount of per-trace state required for online learning to just the last state-action pair, safeguarding the limited memory of the target devices.
We also design our solution to operate as closely as possible to the P4 pipeline to use and learn from per-packet state, but outside of the main packet path to prevent packet stalls.

@@

While P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to great interest in network programmability, it requires similar behaviour between device classes.
This is encoded by the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}---a lower bound on the device capabilities required to implement a P4 dataplane.
%We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
However, many compatible devices have legacies long predating these developments. 
For instance, many-core SoC-based Netronome~\parencite{netronome-smartnic}, NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14,DBLP:conf/fpga/IbanezBMZ19}, and other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo} allow virtually arbitrary programs to be specified and executed.
%This pattern extends to other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo}.
Currently, low port-density devices like these are most likely to have specific general-purpose compute, as they are designed for high-performance offloading and middlebox development.
%Even Intel Tofino ASICs~\parencite{barefoot-intel}, which architecturally mirror the PSA, expose additional matching capabilities and \emph{arithmetic logic unit} (ALU) functions via the Tofino Native Architecture.
Regardless, floating-point operations key to ML workloads are very rarely supported outside of deep learning accelerators (which lack packet switching functionality).
Moreover, the limited memory and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

@@

\emph{Reinforcement learning} (RL) algorithms train an \emph{agent} to choose an optimal sequence of actions in pursuit of a given task~\parencite{RL2E}.
Like most ML methods, RL algorithms use gradient information to update the parameters used to approximate a function with \emph{reward measurements} from the environment.
RL's \emph{Markov Decision Process} (MDP) structure allows online learning of a state-action map in a model-free way, and can step through local minima when needed compared to ML.
In networking, this allows for learning to respond to on-device state or to handle rapidly evolving problems.
RL algorithms are generally agnostic to the policy approximation used, and classical single-step methods are simple---most of the compute work is occupied by the policy.
E.g., single-step, semi-gradient Sarsa~\parencite[pp. \numrange{217}{221}]{RL2E} requires only additions and multiplication to update and learn a policy online.

@@

\subsection{Representations for embedded ML}
PDP hardware universally lacks floating-point arithmetic support, yet the task of bringing ML models to resource-constrained environments is well-studied.
Quantisation and alternative data formats have been suggested to make ML inference feasible on resource- and power-limited platforms, work around hardware constraints, or compute faster and more efficiently.
Lower bit depths reduce memory footprints, and improve throughput in e.g., \emph{bfloat16}~\parencite{bfloat16-blog,DBLP:journals/sigops/XieDMKVZT18}, \emph{hfp8}~\parencite{DBLP:conf/nips/SunCCWVSCZG19}, and other floating point formats~\parencite{DBLP:journals/corr/abs-2007-01530}.
In many cases, accuracy losses are negligible.
Such work goes further still towards integers~\parencite{tensorrt-8bit} or binarisation~\parencite{DBLP:journals/corr/MiyashitaLM16,DBLP:conf/eccv/RastegariORF16,DBLP:journals/corr/KimS16,DBLP:conf/nips/HubaraCSEB16}, sacrificing dynamic range for cheaper arithmetic.
Similar techniques are used to make inference tractable on network hardware, i.e., to bring pre-trained neural networks to the dataplane~\parencite{DBLP:journals/corr/abs-2009-02353,DBLP:conf/sigcomm/SanvitoSB18,DBLP:journals/corr/abs-1801-05731}.

@@

\subsection{Action and Update Computation}\label{sec:action-and-update-computation}
\approachshort{} applies the insights of \textcite{DBLP:journals/firai/TravnikMSP18} to minimise action latency; an action is computed, sent out into the environment, and only then is the underlying policy updated.
Using one of the below strategies chosen at compile time, a state vector is tile coded, converted into action probabilities, and an action is chosen.
This is then written out to the environment as in \cref{sec:agent-environment-communication}.
If online learning is enabled, \approachshort{} then checks an internal hashmap for a previous state-action pair matching the current instruction source, and if found then the policy is updated.
Updates are computed using \emph{single-step semi-gradient Sarsa}~\cite[pp. \numrange{217}{221}]{RL2E}, though modification to support other single-step methods would be trivial.
The new state-action pair is then written into storage.
\approachshort{} can be configured to automatically select values of the input state vector as keys for state and reward storage.
Two firmware models govern how these tasks (action selection, policy updates) are carried out:
\begin{description}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each works sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate to process state vectors, minimising latency. \emph{Minion} threads have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{description}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks to update the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput have different effects on RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem).

@@

\subsection{Agent-Environment Communication}\label{sec:agent-environment-communication}
\approachshort{} uses \emph{multiple-producer/multiple-consumer} (MPMC) channels to communicate with other elements; be they P4 programs on the packet path, or other on-chip analysis and control modules.
Through these, a system \emph{pushes} state vectors, reward measures, and setup packets as inputs, and \emph{pulls} a stream of state-action pairs as outputs.
This allows decisions to be made asynchronously, preventing packet stalling.
The key insight of this mechanism is that on-chip reward or state signals enjoy first-class support in the same manner as packets from the P4 dataplane, allowing agents to act on environmental signals from other on-chip processes or the controller.
As such, \approachshort{} can receive input from P4 \texttt{extern}s or other, dedicated off-path flow state measurement applications.

Our implementation uses platform-specific IPC (\emph{EMEM ring buffers}) with hardware signalling on work arrival to achieve this.
As PDP hardware lacks dynamic memory allocation, P4 pipeline threads request buffers for packet payloads using a shared freelist to enable state or policy data handover.
Packet headers are extracted and parsed using the tooling autogenerated by the P4 pipeline.
We found that this costs a median \SIrange{126}{140}{\nano\second} communication time (local--remote), comparable to message channels in the Rust and Go languages on commodity hardware.

@@

Our implementation exploits the locality of cores in the NFP. 
Policy compute or update tasks are passed between cores using direct \emph{next neighbour} registers, signalling all child threads in response.
Such on-chip signals cost just $\sim$\SI{20}{\nano\second} per relayed message.
This can be factored into the design of additional async off-path functional units (similar to \textcite{DBLP:conf/isca/LiLYCSH19}) on platforms like NetFPGA.
Each core performs atomic adds to a shared preference list and an acknowledgement counter checked by the master thread, implementing our wait-free \emph{ParSa} algorithm (\cref{alg:parsa}).
This is essential for aggregation compared to the use of bounded message buffers, which caused significant head-of-line blocking.
%For instance, additional off-path functional units can serve as workers, like the floating-point adders used by \textcite{DBLP:conf/isca/LiLYCSH19}.
%Rather than general-purpose IPC, \approachshort{} would be hard-wired between internal workers and to the P4$\rightarrow$NetFPGA~\parencite{DBLP:conf/fpga/IbanezBMZ19} actions that interface with it.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and learning parameters to be changed at runtime using at most two control packets.
Design changes are used at the end of learning (moving from online to offline), or when trying to train a new policy for another task from the same vantage point.
Parameter changes allow an online agent to become more (or less) adaptive to new data (i.e., after detecting a changepoint in traffic).
This extends to policy data, which may be imported from a pre-trained model and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, \Coopfw/\Indfw, and maximum policy sizes---these govern core operation or allocated memory.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing more complex problems to be modelled using more dimensions or fine-grained tiles.

@@

Policy sizes are set to those of a real-world DDoS control application~\parencite{DBLP:journals/tnsm/SimpsonRP20}: 20-dim state vectors, a bias tile and 16 full tiling sets (\numproduct{7x1}-dim, \numproduct{8x2}-dim, \numproduct{1x4}-dim), 8 tilings per set, 6 tiles per dimension, and 10 actions.
For context, such input would contain per-flow state (e.g., IATs, rates) which is combined with the last action taken (2-dim tilings) and loads along the ingress-egress path (4-dim).
In \Coopfw{}, this creates \num{129} work items across \num{31} workers.
We choose a larger action count than this work to investigate the performance of more complex agents.

@@

We then vary the amount of worker threads to show how \approachshort{} scales to fit available compute resources on a device.
This is an important aspect for (automated) allocation of compute resources in an intelligent dataplane---particularly when cohabiting with other dataplane programs---and has effects on ahead-of-time work scheduling which we examine later.
This also demonstrates the number of cores needed to achieve a given latency or throughput bound on a policy of representative complexity.
Moreover, to demonstrate how these costs vary as policy complexity increases, we vary the number of total dimensions included in the tiling.

@@

We perform these tests using Pktgen-DPDK~\cite{pktgen-dpdk}, placing an NFP in \emph{MidServer} as the device under test and connecting \emph{HighServer} over a \SI{40}{\giga\bit\per\second} DAC cable as the traffic source via the default firmware.
We perform throughput and loss tests using \num{7}/\num{1} transmit/receive queues at \SI{100}{\percent} send rate for 10 bursts of \SI{30}{\second}, and perform latency tests using \num{1}/\num{1} transmit/receive queue at \SI{10}{\percent} send rate for \num{200000} measurements (sampling at \SI{2000}{\hertz} for \qtyproduct[product-units=single]{10 x 10}{\second}).
This provides maximum throughput in the former case (relying solely on NIC counters for loss counting).
In the latter case, this minimises host resource contention to observe exact latency measurements, have a high enough sample count to detect subtle (aggregate) latency effects, and eliminate \emph{host} receive drops.
DPDK was setup using \qtyproduct[product-units=single]{4 x 1}{\gibi\byte} hugepages.
Sent traffic was comprised of fixed-size \SIrange{64}{1518}{\byte} packets~\cite{rfc2544}.
CPU clock scaling was disabled on \emph{HighServer}.

\fakepara{Resource requirements}
Using the maximum policy size defined above, we investigate how the memory requirements imposed by \approachshort{} vary with the number of dedicated MEs, over and above a base P4 forwarding plane.
We report resource use for \SI{32}{\bit} \Indfw{} and \Coopfw{} agents, with hash tables sized to \num{4096} state-action pairs and \num{16} separate reward values.
This captures the relative cardinality of network RL traces to rewards, i.e., many input flows will map to one or few reward values (i.e., DDoS attack size estimation per egress-AS, queue occupancy in the case of AQM per output port).

@@

By limiting the available workers in software, we show how \Coopfw{}'s policy update time (thus  online throughput---\cref{fig:vary-core}) scales with available cores.
While \Coopfw{} always outperforms the host-based floating point implementations, we observe that there are two distinct crossover points which must be met to overcome our own \Indfw{}; \num{8} workers for online throughput, and \num{3} workers for state-action latency (plot omitted).
Some artefacts of our environment and design choices are visible, such as the addition of new physical cores being more significant than contexts, and the presence of some schedule bottlenecks.
Most importantly however, \Coopfw{}'s resource demand is tunable at compile time to meet the online training rate or action latency required by a task.

@@

%A noteworthy trend is that \SIlist{8;16}{\bit} versions of \approachshort{} consistently underperform compared to \SI{32}{\bit}, except for smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency}).
\SIlist{8;16}{\bit} versions of \approachshort{} underperform compared to \SI{32}{\bit}, except for smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency})---even though our implementation is optimised to read and write policy data in batches.
Because the native register width on the NFP is \SI{32}{\bit}, the compiler must emit extra instructions around arithmetic operations to correctly load and update values.
This matches \SI{32}{\bit} becoming dominant in complex workloads: higher dimension tilings require more arithmetic operations.
Most of the I/O comes \emph{after} this step, causing ALU use to dominate.
This also explains why \SI{32}{\bit} becomes the best choice at different policy complexities for online (\cref{fig:vary-work}, 10 dims) and offline (\cref{fig:vary-work-latency}, 3 dims) agents, where hash table accesses and a \texttt{memcpy} of the state vector fall into the serial portion of the online algorithm.
%Smaller bit depths still give a \SIrange{2}{4}{$\times$} saving in memory for policy storage (enabling more fine-grained or complex policies), in exchange for slightly worse latency and throughput.
To overcome this, we investigated bit-stuffing several values into a single word during atomic writeback (as the platform offers both \SIlist{32;64
}{\bit} atomic addition).
This is analogous to SIMD through clever use of padding bits, but we found that manipulating tiles into the correct format added \SI{10}{\percent} extra overhead.

@@

\fakepara{Resource requirements}
\Cref{tab:resources} shows how \approachshort{} consumes memory as it scales to fit a device's compute resources, compared with a simple P4 forwarding application.
As one program is installed per ME, these results represent the minimum and maximum resource use on a single island (i.e., without replacing P4 workers).
We observe negligible costs on shared EMEM ($\sim$\SI{4}{\mebi\byte}) from state and reward hash tables.
The most significant costs arise due to policy data (\SI{405}{\kibi\byte} shared IMEM, \SI{90}{\kibi\byte} local CTM, \SI{15}{\kibi\byte} local CLS), which can be halved or quartered using \SIlist{16;8}{\bit} quantisation.
This is a high upfront cost on per-island resources (CLS/CTM)---\approachshort{} leaves resources for other off-path dataplane applications, but is fairest from 3 cores onwards.
